---
title: "Lab 6: Exporting Data and Output from R"
subtitle: "Stat 159, Fall 2016, Prof. Sanchez"
date: "October 4, 2016"
output: pdf_document
fontsize: 12pt
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Data Set `mtcars`

To illustrate the different data exporting possibilities, as well as writing 
output features in R, we are going to use the data frame `mtcars` that comes 
in R:

```{r, comment=''}
head(mtcars)
```


## Writing tables 

One common task in most data analysis projects involves exporting data tables.
For this purposes you can use `write.table()` or `write.csv()`.
Here are a couple of examples that export `mtcars` to text files using
different types of field separators. Note that these examples assume you want 
the files in your working directory.

```{r eval = FALSE}
# blank separated (default)
write.table(mtcars, file = 'mtcars.txt', row.names = FALSE)

# tab-separated value
write.table(mtcars, file = 'mtcars.tsv', sep = "\t", row.names = FALSE)

# comma-separated value
write.csv(mtcars, file = 'mtcars.csv', row.names = FALSE)
```


## Sending output with `cat()`

You can use `cat()` to concatenate and print information to a file. 
For instance, say you are interested in some descriptive statistics about 
the column `mpg` (miles-per-gallon):
```{r eval = FALSE}
# summary statistics of mpg
min(mtcars$mpg)
max(mtcars$mpg)
median(mtcars$mpg)
mean(mtcars$mpg)
sd(mtcars$mpg)
```

The goal is to generate a file `mpg-statistics.txt` with the following 
contents:

```
Miles per gallon summary statistics

Minimum: 10.4
Maximum: 33.9
Median : 19.2
Mean   : 20.09 
Std Dev: 6.02 
```

Here's one way to start:
```{r eval = FALSE}
# summary statistics of mpg
mpg_min <- min(mtcars$mpg)
mpg_max <- max(mtcars$mpg)
mpg_med <- median(mtcars$mpg)
mpg_avg <- mean(mtcars$mpg)
mpg_sd <- sd(mtcars$mpg)

# name of output file
outfile <- "mpg-statistics.txt"

# first line of the file
cat("Miles per gallon summary statistics\n\n", file = outfile)
# subsequent lines appended to the output file
cat("Minimum:", mpg_min, "\n", file = outfile, append = TRUE)
cat("Maximum:", mpg_max, "\n", file = outfile, append = TRUE)
cat("Median :", mpg_med, "\n", file = outfile, append = TRUE)
cat("Mean   :", mpg_avg, "\n", file = outfile, append = TRUE)
cat("Std Dev:", mpg_sd, "\n", file = outfile, append = TRUE)
```

To make it "prettier" you may consider using `sprintf()`
```{r eval = FALSE}
sprintf('Minimum: %s', mpg_min)
```

Now let's re-export the lines:
```{r eval=FALSE}
cat("Miles per gallon summary statistics\n\n", file = outfile)
cat(sprintf('Minimum: %0.2f', mpg_min), "\n", file = outfile, append = TRUE)
cat(sprintf('Maximum: %0.2f', mpg_max), "\n", file = outfile, append = TRUE)
cat(sprintf('Median : %0.2f', mpg_med), "\n", file = outfile, append = TRUE)
cat(sprintf('Mean   : %0.2f', mpg_avg), "\n", file = outfile, append = TRUE)
cat(sprintf('Std Dev: %0.2f', mpg_sd), "\n", file = outfile, append = TRUE)
```

__Your turn:__
How would you avoid writing that many calls to `cat()`?


## Sending R output to a file with `sink()`

Another interesting function is `sink()`. This function is very useful when
you want to export R output as is displayed in the R console. For example, 
consider the output from `summary()`
```{r, comment=''}
summary(mtcars$mpg)
```

You could assign the output of `summary(mtcars$mpg)` to an object `mpg_summary` 
and then try `writeLines()` to export the results to a file `mpg-summary.txt`, 
but you won't keep the same format of R:
```{r eval = FALSE}
mpg_summary <- summary(mtcars$mpg)
writeLines(mpg_summary, con = "mpg-summary.txt")
```

To be able to keep the same output display of R, you must use `sink()`. This
function will __divert__ or redirect R output to the specified file:
```{r eval=FALSE}
# sink output
sink(file = "mpg-stats2.txt")
# summary statistics of mpg
summary(mtcars$mpg)
# stops diverting output
sink()
```


__Your turn:__
Use `sink()` to send the output from running a linear regression of `hp` on
`mph` with the function `lm()`. Also export the results from
using `summary()` on the regression object. And/or try running a t-test
between `mpg` and `hp` with `t.test()`.


## Exporting tables with `xtable()`

Another interesting tool to export tables in LaTeX or HTML formats is provided
by the R package `"xtable"` and its main function `xtable()`.

```{r eval = FALSE}
library(xtable)

# linear regression
reg <- lm(hp ~ mpg, data = mtcars)

# create xtable and export it
reg_table <- xtable(reg)
print(reg_table, type = "latex", file = "reg-table.tex")
print(reg_table, type = "html", file = "reg-table.html")
```


## R's Binary Data

R also allows you to save objects in R's binary format with the functions 
`save()` and `save.image()`. It is customary to use the `.RData` extension for 
the files created by `save()` and `save.image()`. You may also encounter 
users specifying the old extension `.rda` or some other variation.

You can use `save()` to save specific objects from your current session.
For example, here is how to save the data frame `mtcars` to your working dir:
```{r eval=FALSE}
save(mtcars, file = 'mtcars.RData')
```

The difference between `save()` and `save.image()` is that the latter saves
all the objects in your current session. This is actually the function that is
run behind the scenes everytime you quit R and accept to save the so-called
_workspace image_.

You can share `mtcars.RData` with any other R user, regardless of the 
operating system that they use. To read in binary R files, use `load()`.

__Your turn:__ 
Subset the data set `mtcars` for cars with automatic transmission, `am == 1`, 
and export the resulting data using both `write.table()` and `save()`. 
Compare the size of the produced files.



