---
title: "Lab 7: Unit tests with testthat"
author: "Stat 159, Fall 2016, Prof. Sanchez"
date: "October 10, 2016"
output: pdf_document
urlcolor: blue
---

> ### Learning Objectives
>
> * Introduction to the R package "testthat"
> * Write simple functions and their unit tests
> * Test your code


## R package `"testthat"`

`"testthat"` is one of the packages in R that helps you write tests for your functions. One of the main references is the paper 
_testthat: Get Started with Testing_ by Hadley Wickham (see link below). 
This paper clearly describes the philisoply and workflow of `"testthat"`. But
keep in mind that since the introduction of the package, many more functions 
haven been added to it.

[https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf](https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf)


## About `"testthat"`

- `"testthat"` provides a testing framework for R that is easy to learn and use
- `"testthat"` has a hierarchical structure made up of:
    + expectations
    + tests
    + contexts
- A __context__ involves __tests__ formed by groups of __expectations__
- Each structure has associated functions:
    + `expect_that()` for expectations
    + `test_that()` for groups of tests
    + `context()` for contexts


### List of common expectation functions

| Function                  | Description                             |
|---------------------------|-----------------------------------------|
| `expect_true(x)`          | expects that `x` is `TRUE`              |
| `expect_false(x)`         | expects that `x` is `FALSE`             |
| `expect_null(x)`          | expects that `x` is `NULL`              |
| `expect_type(x)`          | expects that `x` is of type `y`         |
| `expect_is(x, y)`         | expects that `x` is of class `y`        |
| `expect_length(x, y)`     | expects that `x` is of length `y`       |
| `expect_equal(x, y)`      | expects that `x` is equal to `y`        |
| `expect_equivalent(x, y)` | expects that `x` is equivalent to `y`   |
| `expect_identical(x, y)`  | expects that `x` is identical to `y`    |
| `expect_lt(x, y)`         | expects that `x` is less than `y`       |
| `expect_gt(x, y)`         | expects that `x` is greater than `y`    |
| `expect_lte(x, y)`        | expects that `x` is less than or equal to `y` |
| `expect_gte(x, y)`        | expects that `x` is greater than or equal `y` |
| `expect_named(x)`         | expects that `x` has names `y`          |
| `expect_matches(x, y)`    | expects that `x` matches `y` (regex)    |
| `expect_message(x, y)`    | expects that `x` gives message `y`      |
| `expect_warning(x, y)`    | expects that `x` gives warning `y`      |
| `expect_error(x, y)`      | expects that `x` throws error `y`       |

-----

## Practice

- To start the practice, create a new directory, e.g. `test-that`
- `cd` into the directory `test-that/`
- Create two subdirectories: `functions` and `tests`
- Sample content is in the folder `test-that/` of this lab


## Functions

Let's start with a couple of basic functions: `range_value()` and 
`missing_values()`:

```{r}
range_value <- function(x) {
  max(x) - min(x)
}
```

- _description:_ computes the range of a numeric vector (i.e. max - min)
- _input:_ a numeric vector
- _output:_ the range value (max - min)


```{r}
missing_values <- function(x) {
  sum(is.na(x))
}
```

- _description:_ computes the number of missing values
- _input:_ a numeric vector
- _output:_ the number of missing values


Write these functions in files `range-value.R` and `missing-values.R` 
inside `functions/`


## Tests

In the folder `tests/`, create a file `tests.R` that includes tests for
`range_value()` and `missing_values()`.

To write the unit tests, we are going to consider the following testing vectors:

- `x <- c(1, 2, 3, 4, 5)`
- `y <- c(1, 2, 3, 4, NA)`
- `z <- c(TRUE, FALSE, TRUE)`
- `w <- letters[1:5]`

The typical structure of the tests has the following form:

```{r eval = FALSE}
# load the source code of the functions to be tested
source("../functions/functions.R")

# context with one test that groups expectations
context("Test for range value") 

test_that("range works as expected" {
  x <- c(1, 2, 3, 4, 5)
  
  expect_equal(range_value(x), 4)
  expect_length(range_value(x), 1)
  expect_type(range_value(x), 'double')
})
```

- use `context()` to describe what the test are about
- use `test_that()` to group expectations:
    + output equal to 4
    + output of length one
    + output of type `double`
- to run the tests from the R console, use the function `test_file()`

```{r eval = FALSE}
# assuming that your working directory is "test-that/"
# from the R console
library(testthat)
test_file("tests/tests.R")
```


### Your Turn

Write more tests---`test_that()`---to test `range_value()` with the rest of
the testing vectors `y`, `z`, `w`:

- Using `y`, write expectations for:
    + output of length one
    + output is `NA_real_`
- Using `z`, write expectations for:
    + output of length one
    + output of type `integer`
    + output equal to `1L`
- Using `w`, write expectations for:
    + throws an error


-----

## Missing Values

Now consider the function `missing_value()`.
Write a context for testing the following expectations:

- output of length one
- output of type `double`
- output greater than or equal to zero


## Improving `range_value()`

Modify the function `range_value()` to include an argument `na.rm` that
takes a logical value indicating whether missing values should be removed
before computing the range.

Adapt the tests for the improved `range_value()` function, especially when
using the testing vector `y <- c(1, 2, 3, 4, NA)`.


## Extra Challenges

Try writing the following functions and come up with unit tests:

- `center_measures()`
    - _description:_ computes measures of center such as Median and Mean
    - _input:_ a numeric vector
    - _output:_ a numeric vector with median and mean

- `spread_measures()`
    - _description:_ computes measures of spread such as Range, IQR, Std Dev
    - _input:_ a numeric vector
    - _output:_ a numeric vector with range, iqr, and stdev

- `descriptive_stats()`
    - _description:_ computes descriptive statistics
    - _input:_ a numeric vector
    - _output:_ a numeric vector with median, mean, range, iqr, stdev, and
    number of missing values.

-----
